Early Observations of the Type Ia Supernova \abc:\\ Evidence for Strong Ejecta Mixing and/or Interaction with Diffuse Material

### 2017-08-21 doc/ms.tex
Update text following Jesper's last comments and spell check

Interaction is not limited to systems with a non-degenerate companion, however, as WDs enshoruded in diffuse material following a binary merger (e.g., \citealt{2015MNRAS.447.2803L}) or expanded due to a pre-explosion pulsation can give rise to ejecta-interaction signatures (e.g., \citealt{2014MNRAS.441..532D}). Models of this scenario naturally produce \ion{C}{2} absorption that is comparable in strength to \ion{Si}{2} in the days after explosion \citep{2014MNRAS.441..532D}, as was observed in SN\,2013dy \citep{2013ApJ...778L..15Z} and SN\,2017cbv \citep{2017arXiv170608990H}.

@@

The time of first light, defined as the moment when appreciable heating from $^{56}$Ni decay has first diffused to the photosphere, for SNe is usually estimated by 
extrapolating early-phase light curves to determine when the 
SN flux is equal to 0. Assuming an ideal, expanding fireball with 
constant temperature, \citet{1982ApJ...253..785A} derives that $f 
\propto t^2$, where $f$ is the SN flux and $t$ is the time since 
explosion. Despite the basic assumption of a constant temperature at 
early times, multiple studies have found that the early emission 
from Type Ia SNe can be described as a power law in time, with 
power-law index consistent with 2, i.e. $f \propto t^2$ (e.g., \citealt{2006AJ....132.1707C, 2010ApJ...712..350H, 2011MNRAS.416.2607G}).\footnote{Many of the studies that find $f \propto t^2$ sample SNe Ia light curves at a phase that is closer to $T_{B,\mathrm{max}}$ than the initial observations of \abc\ discussed here.}

@@

The potential dark phase following shock breakout, where SNe Ia are not luminous because the energy deposition from radioactive decay has not had sufficient time to diffuse to the photosphere, means that the explosion time $t_\mathrm{exp} \ne t_0$ in some cases. \citet{2014ApJ...784...85P} suggest that measurements of the photospheric velocity be used to determine $t_\mathrm{exp}$ given that the ejecta begin expanding from the moment of explosion.  Assuming a constant opacity in the ejecta, \citeauthor{2014ApJ...784...85P} 
find that the photospheric velocity evolves as
$v_\mathrm{ph}\propto(t-t_\mathrm{exp})^{-0.22}$. Numerical experiments by \citet{2016ApJ...826...96P} find that the consatnt-opacity assumption strongly depends on the amount of $^{56}$Ni mixing in the SN ejecta (for a more detailed discussion of $^{56}$Ni mixing see \S\ref{sec:Ni_mixing}). As a result, the adoption of a $t^{-0.22}$ power-law model may not be valid for all SNe Ia. Nevertheless, we proceed on the assumption that \abc\ experienced strong $^{56}$Ni mixing, corresponding to the models that are best approximated as a $t^{-0.22}$ power law. We do this in part to compare with previous studies, though we caution that the inferred value of $t_{exp}$ is subject to uncertainties related to ejecta mixing.

@@

\begin{figure*}[!thb]
  \centering
  \includegraphics[width=0.48\textwidth]{Chi2.pdf}
  \includegraphics[width=0.48\textwidth]{VelocityPlot.pdf}
  \caption{Constraints on $t_{exp}$ from fitting the velocity
    evolution of \ion{Si}{2}.
    \textit{Left panel:} the dashed, solid and dash-dotted curves 
    show $\chi^2$ for fitting power laws with
    indices $-0.20$, $-0.22$ and $-0.24$, respectively. The blue
    vertical line and the orange shaded region indicate $t_0$ and 
    its 95\% confidence interval from Section
    \ref{sec:lc_fit}, respectively.
    \textit{Right panel:} Observed \ion{Si}{2}\,$\lambda$6355
    velocities (blue circles) and the best-fit power-law model
     with an index of $-0.22$ (dashed line). Additionally, the 
     measured velocites of \ion{C}{2}\,$\lambda\lambda$6580, 7234 
     are shown. Typical uncertainties are smaller than the size of 
     the points. \textit{Right inset}: Evolution of the 
     pseudo-equivalent width of \ion{Si}{2}\,$\lambda$6355, and 
     \ion{C}{2}\,$\lambda\lambda$6580, 7234 in the $\sim$7~d    
     following explosion.}
  \label{fig:velocity_t_exp}
\end{figure*}

We fit the measured velocities of \ion{Si}{2}\,$\lambda$6355 to the
$v\propto(t-t_{exp})^{-0.22}$ model by minimizing the $\chi^2$ value and find the best-fit explosion time relative to $T_{B,\mathrm{max}}$ in the SN rest frame to be $t_{exp} = -17.45 \pm ^{0.14}_{0.16}\,\textrm{d}$, where the uncertainties represent the 95\% confidence interval (Figure \ref{fig:velocity_t_exp}). Following the analysis in \citet{2014ApJ...784...85P}, we additionally alter the power-law index to $-0.20$ and $-0.24$ to examine the sensitivety of the result on the assumed power-law index. We find that this variation in the power-law index results in inferred explosion times that vary by $\approx$1 d (Figure \ref{fig:velocity_t_exp}). Given the analytical approximation that $v \propto t^{-0.22}$, we adopt $t_{exp} = -17.5 \pm 0.5 \; \mathrm{d}$, where the uncertainty reflects possible variations in the power-lax index (see \citealt{2014ApJ...784...85P}).

@@

Relative to the nearby, normal SN\,2011fe, we have identified several distinct characteristics of the early evolution of \abc, including: (i) a near-linear photometric rise, (ii) a qualitatively short dark phase, assuming $v_\mathrm{ph} \propto t^{-0.22}$, (iii) the presence of strong \ion{C}{2} absorption, and (iv) blue and non-evolving $(B - V)_0$ color in the week after explosion. While most SNe Ia are powered purely by radioactive decay, the observed radiation shortly after explosion can also include contributions 
from SN shock cooling or the collision of the SN ejecta with a non-degenerate companion or diffuse material. Here we consider those scenarios as a possible explanation for the early behavior of \abc.

@@

Figure~\ref{fig:SN-companion} additionally shows that a companion at $a \approx 18 \times 10^{11} \, \mathrm{cm}$ provides a good match to the initial optical rise, if $t_\mathrm{exp} \approx t_0 + 0.1 \; \mathrm{d}$. Models with $a \gtrsim 10^{12} \, \mathrm{cm}$, which can explain the initial $g_\mathrm{PTF}$ rise, signiticantly overpredict the observed UV flux, however. There is no choice of $a$ capable of replicating the early rise of \abc\ without also overpredicting the observed UV flux.

@@

Having examined other possibilities we now consider whether the unusual properties of \abc\ can be explained simply by invoking strong mixing in the SN ejecta. Strong mixing can lead to a short dark phase. Here we additionally examine whether strong mixing can explain the rapid, near-linear rise in the light curve.

@@

In \S\ref{sec:first_light} we compared our early observations of \abc\ to SN\,2011fe, which has the most comprehensive observations of the young SNe Ia sample. Given its normal spectroscopic and photometric evolution, SN\,2011fe has been adopted as a standard for the early evolution of SNe Ia in many studies. While a detailed quantitative analysis is beyond the scope of this study, a qualitative examination of very young SNe Ia that otherwise exhibit normal spectra and evolution at and post-peak\footnote{This definition excludes iPTF\,14atg, which was shown to be subluminous with SN\,2002es-like spectra \citep{2015Natur.521..328C}.} reveals considerable diversity. In other words, at early times SN\,2011fe may not be the norm.

@@

In the case of SN\,2012cg and SN\,2017cbv it has been argued that the early blue optical colors are indicative of interaction between the SN ejecta and a binary companion \citep{2016ApJ...820...92M,2017arXiv170608990H}. SN\,2017cbv is particularly remarkable in that the observations presented in \citet{2017arXiv170608990H} show a clearly resolved bump in the $U$, $B$, and $g'$ bands in the $\sim$5 d after explosion. In \citeauthor{2017arXiv170608990H} it is found that the bump can be explained via the combination of ejecta-companion interaction and the normal evolution of a SN Ia. A challenge for this model, similar to \abc\ (see \S\ref{sec:companion}), is that it significantly overpredicts the UV brightness of the SN compared to what is observed. Indeed, in the case of SNe\,2012cg, 2017cbv, and \abc\ the $\mathrm{UV} - \mathrm{optical}$ colors are significantly redder than those observed in iPTF\,14atg. It is argued in \citet{2017arXiv170608990H} that several model assumptions, including: (i) ideal blackbody emission, (ii) a constant opacity, (iii) a simple power-law density profile for the ejecta, and (iv) spherical symmetry, may be incorrect, which could reconcile the discrepancy with the UV observations. Above, we argued for circumstellar interaction and strong $^{56}$Ni mixing as a possible explanation for \abc, and indeed \citet{2017arXiv170608990H} consider these possibilities for SN\,2017cbv as well. Seperately, several arguments against companion interaction for SN\,2012cg are presented in \citet{2016arXiv161007601S}.

Ultimately, there are arguments in favor of and against each of the possibilties to explain the early emission from SNe Ia. Moving forward, more detailed models and simulations are needed to properly explain the observed diversity. No matter the correct explanation for the early behavior of SNe\,2013dy, 2017cbv, and \abc, the strong similarities between these events suggests that they may reflect a common physical origin.

@@

We are deeply grateful to R.~C.~Thomas for induldging a slew of questions regarding \ion{C}{2} in SNe. Similarly, it is our pleasure to buy a beer for R.~Amanullah and U.~Feindt for discussions regarding SALT2.

@@

The Intermediate Palomar Transient Factory project is a scientific collaboration among the California Institute of Technology, Los Alamos National Laboratory, the University of Wisconsin, Milwaukee, the Oskar Klein Center, the Weizmann Institute of Science, the TANGO Program of the University System of Taiwan, and the Kavli Institute for the Physics and Mathematics of the Universe. This work was supported by the GROWTH project funded by the National Science Foundation under Grant No 1545949. Part of this research was carried out at the Jet Propulsion Laboratory, California Institute of Technology, under a contract with the NASA. This work makes use of observations from the LCO network. These results made use of the Discovery Channel Telescope at Lowell Observatory. Lowell is a private, non-profit institution dedicated to astrophysical research and public appreciation of astronomy and operates the DCT in partnership with Boston University, the University of Maryland, the University of Toledo, Northern Arizona University and Yale University.  The upgrade of the DeVeny optical spectrograph has been funded by a generous grant from John and Ginger Giovale.

>>>>

Interaction is not limited to systems with a non-degenerate companion, however, as WDs enshrouded in diffuse material following a binary merger (e.g., \citealt{2015MNRAS.447.2803L}) or expanded due to a pre-explosion pulsation can give rise to ejecta-interaction signatures (e.g., \citealt{2014MNRAS.441..532D}). Models of this scenario naturally produce \ion{C}{2} absorption that is comparable in strength to \ion{Si}{2} in the days after explosion \citep{2014MNRAS.441..532D}, as was observed in SN\,2013dy \citep{2013ApJ...778L..15Z} and SN\,2017cbv \citep{2017arXiv170608990H}.

@@

The time of first light, interpreted in this study as the moment when appreciable heating from $^{56}$Ni decay has first diffused to the photosphere, for SNe is usually estimated by 
extrapolating early-phase light curves to determine when the 
SN flux is equal to 0. Assuming an ideal, expanding fireball with 
constant temperature, \citet{1982ApJ...253..785A} derives that $f 
\propto t^2$, where $f$ is the SN flux and $t$ is the time since 
explosion. Despite the basic assumption of a constant temperature at 
early times, multiple studies have found that the early emission 
from Type Ia SNe can be described as a power law in time, with 
power-law index consistent with 2, i.e. $f \propto t^2$ (e.g., \citealt{2006AJ....132.1707C, 2010ApJ...712..350H, 2011MNRAS.416.2607G}).\footnote{Many of the studies that find $f \propto t^2$ sample SNe Ia light curves at a phase that is closer to $T_{B,\mathrm{max}}$ than the initial observations of \abc\ discussed here.}

@@

The potential dark phase following shock breakout, where SNe Ia are not luminous because the energy deposition from radioactive decay has not had sufficient time to diffuse to the photosphere, means that the explosion time $t_\mathrm{exp} \ne t_0$ in some cases. \citet{2014ApJ...784...85P} suggest that measurements of the photospheric velocity be used to determine $t_\mathrm{exp}$ given that the ejecta begin expanding from the moment of explosion.  Assuming a constant opacity in the ejecta, \citeauthor{2014ApJ...784...85P} 
find that the photospheric velocity evolves as
$v_\mathrm{ph}\propto(t-t_\mathrm{exp})^{-0.22}$. Numerical experiments by \citet{2016ApJ...826...96P} find that the constant-opacity assumption strongly depends on the amount of $^{56}$Ni mixing in the SN ejecta (for a more detailed discussion of $^{56}$Ni mixing see \S\ref{sec:Ni_mixing}). As a result, the adoption of a $t^{-0.22}$ power-law model may not be valid for all SNe Ia. Nevertheless, we proceed on the assumption that \abc\ experienced strong $^{56}$Ni mixing, corresponding to the models that are best approximated as a $t^{-0.22}$ power law. We do this in part to compare with previous studies, though we caution that the inferred value of $t_{exp}$ is subject to uncertainties related to ejecta mixing.

@@

\begin{figure*}[!thb]
  \centering
  \includegraphics[width=0.48\textwidth]{Chi2.pdf}
  \includegraphics[width=0.48\textwidth]{VelocityPlot.pdf}
  \caption{Constraints on $t_{exp}$ from fitting the velocity
    evolution of \ion{Si}{2}.
    \textit{Left panel:} the dashed, solid and dash-dotted curves 
    show $\chi^2$ for fitting power laws with
    indices $-0.20$, $-0.22$ and $-0.24$, respectively. The blue
    vertical line and the orange shaded region indicate $t_0$ and 
    its 95\% confidence interval from Section
    \ref{sec:lc_fit}, respectively.
    \textit{Right panel:} Observed \ion{Si}{2}\,$\lambda$6355
    velocities (blue circles) and the best-fit power-law model
     with an index of $-0.22$ (dashed line). Additionally, the 
     measured velocities of \ion{C}{2}\,$\lambda\lambda$6580, 7234 
     are shown. Typical uncertainties are smaller than the size of 
     the points. \textit{Right inset}: Evolution of the 
     pseudo-equivalent width of \ion{Si}{2}\,$\lambda$6355, and 
     \ion{C}{2}\,$\lambda\lambda$6580, 7234 in the $\sim$7~d    
     following explosion.}
  \label{fig:velocity_t_exp}
\end{figure*}

We fit the measured velocities of \ion{Si}{2}\,$\lambda$6355 to the
$v\propto(t-t_{exp})^{-0.22}$ model by minimizing the $\chi^2$ value and find the best-fit explosion time relative to $T_{B,\mathrm{max}}$ in the SN rest frame to be $t_{exp} = -17.45 \pm ^{0.14}_{0.16}\,\textrm{d}$, where the uncertainties represent the 95\% confidence interval (Figure \ref{fig:velocity_t_exp}). Following the analysis in \citet{2014ApJ...784...85P}, we additionally alter the power-law index to $-0.20$ and $-0.24$ to examine the sensitivity of the result on the assumed power-law index. We find that this variation in the power-law index results in a change of $t_\mathrm{exp}$ of $\approx\pm$0.5\,d (Figure \ref{fig:velocity_t_exp}). Given the analytical approximation that $v \propto t^{-0.22}$, we adopt $t_{exp} = -17.5 \pm 0.5 \; \mathrm{d}$, where the uncertainty reflects possible variations in the power-lax index (see \citealt{2014ApJ...784...85P}).

@@

Relative to the nearby, normal SN\,2011fe, we have identified several distinct characteristics of the early evolution of \abc, including: (i) a near-linear photometric rise, (ii) a qualitatively short, or possibly absent, dark phase, assuming $v_\mathrm{ph} \propto t^{-0.22}$, (iii) the presence of strong \ion{C}{2} absorption, and (iv) blue and non-evolving $(B - V)_0$ color in the week after explosion. While most SNe Ia are powered purely by radioactive decay, the observed radiation shortly after explosion can also include contributions 
from SN shock cooling or the collision of the SN ejecta with a non-degenerate companion or diffuse material. Here we consider those scenarios as a possible explanation for the early behavior of \abc.

@@

Figure~\ref{fig:SN-companion} additionally shows that a companion at $a \approx 18 \times 10^{11} \, \mathrm{cm}$ provides a good match to the initial optical rise, if $t_\mathrm{exp} \approx t_0 + 0.1 \; \mathrm{d}$. Models with $a \gtrsim 10^{12} \, \mathrm{cm}$, which can explain the initial $g_\mathrm{PTF}$ rise, significantly overpredict the observed UV flux, however. There is no choice of $a$ capable of replicating the early rise of \abc\ without also overpredicting the observed UV flux.

@@

Having examined other possibilities we now consider whether the unusual properties of \abc\ can be explained simply by invoking strong mixing in the SN ejecta. Strong mixing leads to a shorter dark phase than would otherwise be observed. Here we additionally examine whether strong mixing can explain the rapid, near-linear rise in the light curve.

@@

In \S\ref{sec:first_light} we compared our early observations of \abc\ to SN\,2011fe, which has the most comprehensive observations of the young SN Ia sample. Given its normal spectroscopic and photometric evolution, SN\,2011fe has been adopted as a standard for the early evolution of SNe Ia in many studies. While a detailed quantitative analysis is beyond the scope of this study, a qualitative examination of very young SNe Ia that otherwise exhibit normal spectra and evolution at and post-peak\footnote{This definition excludes iPTF\,14atg, which was shown to be subluminous with SN\,2002es-like spectra \citep{2015Natur.521..328C}.} reveals considerable diversity. In other words, at early times SN\,2011fe may not be the norm.

@@

In the case of SN\,2012cg and SN\,2017cbv it has been argued that the early blue optical colors are indicative of interaction between the SN ejecta and a binary companion \citep{2016ApJ...820...92M,2017arXiv170608990H}. SN\,2017cbv is particularly remarkable in that the observations presented in \citet{2017arXiv170608990H} show a clearly resolved bump in the $U$, $B$, and $g'$ bands in the $\sim$5 d after explosion. In \citeauthor{2017arXiv170608990H} it is found that the bump can be explained via the combination of ejecta-companion interaction and the normal evolution of a SN Ia. A challenge for this model, similar to \abc\ (see \S\ref{sec:companion}), is that it significantly overpredicts the UV brightness of the SN compared to what is observed. Indeed, in the case of SNe\,2012cg, 2017cbv, and \abc\ the $\mathrm{UV} - \mathrm{optical}$ colors are significantly redder than those observed in iPTF\,14atg. It is argued in \citet{2017arXiv170608990H} that several model assumptions, including: (i) ideal blackbody emission, (ii) a constant opacity, (iii) a simple power-law density profile for the ejecta, and (iv) spherical symmetry, may be incorrect, which could reconcile the discrepancy with the UV observations. Above, we argued for circumstellar interaction and strong $^{56}$Ni mixing as a possible explanation for \abc, and indeed \citet{2017arXiv170608990H} consider these possibilities for SN\,2017cbv as well. Separately, several arguments against companion interaction for SN\,2012cg are presented in \citet{2016arXiv161007601S}.

Ultimately, there are arguments in favor of and against each of the possibilities to explain the early emission from SNe Ia. Moving forward, more detailed models and simulations are needed to properly explain the observed diversity. No matter the correct explanation for the early behavior of SNe\,2013dy, 2017cbv, and \abc, the strong similarities between these events suggests that they may reflect a common physical origin.

@@

We are deeply grateful to R.~C.~Thomas for indulging a slew of questions regarding \ion{C}{2} in SNe. Similarly, it is our pleasure to buy a beer for R.~Amanullah and U.~Feindt for discussions regarding SALT2.

@@

The Intermediate Palomar Transient Factory project is a scientific collaboration among the California Institute of Technology, Los Alamos National Laboratory, the University of Wisconsin, Milwaukee, the Oskar Klein Center, the Weizmann Institute of Science, the TANGO Program of the University System of Taiwan, and the Kavli Institute for the Physics and Mathematics of the Universe. This work was supported by the GROWTH project funded by the National Science Foundation under Grant No 1545949. Part of this research was carried out at the Jet Propulsion Laboratory, California Institute of Technology, under a contract with the NASA. This work makes use of observations from the LCO network. These results made use of the Discovery Channel Telescope at Lowell Observatory. Lowell is a private, non-profit institution dedicated to astrophysical research and public appreciation of astronomy and operates the DCT in partnership with Boston University, the University of Maryland, the University of Toledo, Northern Arizona University and Yale University.  The upgrade of the DeVeny optical spectrograph has been funded by a generous grant from John and Ginger Giovale. Based on observations made with the Nordic Optical Telescope, operated by the Nordic Optical Telescope Scientific Association at the Observatorio del Roque de los Muchachos, La Palma, Spain, of the Instituto de Astrofisica de Canarias.
